\large{\bf Abnormal quantitative genetics? }

### 2023-07-03 writeup.tex
independence bits

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\mL}{\mathcal{L}}
\newcommand{\ip}[2]{\left\langle #1, #2 \right\rangle}
\newcommand{\ie}{\textit{i.e.,}\,}
\newcommand{\eg}{\textit{e.g.,}\,}
\newcommand{\nb}{\textit{n.b.,}\, }
\newcommand{\defn}{:=}

@@

% 1. *Intro*: Motivation about how many results about quant gen and methods assume Gaussian distributions.
%     Mention Levy walk macroevolution papers.
%     Point out that usual quant gen has *no* sweeps, while stable distributions would have sweeps (since some s would be bigger than 1/N).
%     Motivate with mixture model of effect sizes.
%     If stable distributions are a better model, what does that tell us? We can better understand the *noise* in evolution, i.e., predictability and saltation-ness.
%     (PETER)
% 
% 2. Motivation: empirical SNPs distributions in real data (PETER)
% 
% 3. Introducing infinitesimal model (in this case) (TODD)
% 
% 4. Proof following BEV in stable case (TODD)
% 
% 5. Theoretical results on stationary distributions (TODD)
% 
% 6. Simulations (PETER)

\title{\large{\bf
    Quantitative genetics with stable laws
}}

@@

Much of the work to give a quantitative understanding to the observation that
offspring of sexual organisms resemble their parent,
has focused on \emph{quantitative genetics},
which is interestingly intermediate between a mechanistic model of genetics
and a purely phenomenological model.
The action of natural selection on genetic variants in the absence of genetic variation
is relatively well-understood (cite ewens),
but tractable mathematical models of evolution for recombining species
are elusive,
at least with some results of when the model might be a good description of reality.
For instance, a great deal of conceptual understanding and even quantitative predictions
have been made assuming that the distribution of trait values in a population is Gaussian
(cite Lande etcetera)
despite the absence of a plausible model that would make this true (cite turelli and barton).
One of the oldest and most concrete models of heritable trait evolution is the \emph{infinitesimal model}
which assumes a simple, explicit form for how an offspring's trait is inherited from their parents':
it is Normally distributed, with mean equal to the average of the parents' values.
Despite the inherent plausibility of this model,
(thanks to the Central Limit Theorem),
it was not until \citet{barton2017infinitesimal} that conditions under which this would be a good approximation
were made explicit,
through a formal convergence proof.

@@

The infinitesimal model gains a great deal of generality by appealing to the Central Limit Theorem:
as shown by \citet{barton2017infinitesimal},
the distribution of segregation noise will be Normal, regardless of most of the details,
thanks to the universality of adding up a great many very small and independent things.
However, as is well known to probabilists,
there is a wider class of Central Limit Theorems,
which describe the distributions of sums of a great many independent and exchangeable things:
the \emph{stable laws}.
Roughly speaking, these say that if we add together $n$ independent copies of a random variable $X$
for which $\P\{ X > t \} \propto t^{-\alpha}$ for some $0 < \alpha < 2$,
then -- almost regardless of the details -- the result, scaled by $n^{1/\alpha}$,
is well-approximated by a universal form, the $\alpha$-stable distribution (and becomes exact, as $n \to \infty$).
In such ensembles, sinle large entries are important:
concretely, the largest of the $n$ copies will be of order $n^{1/\alpha}$,
so that e.g., for $\alpha=1$ a positive proportion of the sum of $n$ entries will be contributed by a single one.

This suggests exploring whether stable distributions might reasonably stand in for the Normal
in the infinitesimal model.
In this paper, we make rather preliminary explorations in that direction:
Does the universality shown by \citet{barton2017infinitesimal}
carry over when the distribution of effect sizes falls in the domain of attraction of a stable law?
What can be said about equilibrium trait distributions in this case?
And, is there any biological evidence that this theory might be relevant?

@@

>>>>

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\mL}{\mathcal{L}}
\newcommand{\ip}[2]{\left\langle #1, #2 \right\rangle}
\newcommand{\ie}{\textit{i.e.,}\,}
\newcommand{\eg}{\textit{e.g.,}\,}
\newcommand{\nb}{\textit{n.b.,}\, }
\newcommand{\defn}{:=}
\newcommand{\var}{\mathop{\mbox{Var}}}
\newcommand{\cov}{\mathop{\mbox{Cov}}}

@@

% Intro:
%  - the infinitesimal model is useful because it is a "trait-only" model: it allows for predictions and modeling without knowing the underlying genetics.
%  - Its key properties are: (a) independence of seg noise, and (b) Gaussian-ness.
%  - It is basically a CLT, and in particular, (b) derives from a the CLT deriving from similar effect sizes.
%  - Motivating Q: there are other CLTs; can we generalize (b) to non-Gaussian effects, while keeping independence of seg variation?
% 
% Data motivation:
%  First, to see if this is worth asking we look at some data
%  ... and find that yeah, maybe effect sizes don't fall into the Gaussian domain of attraction.
% 
% First answer to the motivating question:  "No"
% Non-Gaussian effects (sometimes) imply nonindependence of segregation noise:
%   In at least some cases, natural non-gaussian models produce nonindependence:
%   1. all loci are uniformly small except there's one big one
%   2. loci are Cauchy
%   3. (is there a simple statement we can make more generally?)
% 
% Second answer, weaker but also "no":
%  - Observation: Let A and B be the contributions to parent phenotype, with A inherited by the offspring and B not. If A and B are independent, and seg noise is independent of parent phenotype, then A and B are Gaussian.
%  - When would A and B be asymptotically independent? One case is the rare-mutation limit that obtains a "thinned Poisson process" limit for A and B (I think: if all allele frequencies are at frequency O(1/M)?)
%  - (I have other thoughts to be sorted out here)
%  - Open Q (maybe?): is there any non-Gaussian model where segregation noise is independent of parental traits?
% 
% Simulations: let's see how this stuff works in practice
%  - neutral simulations
%  - also, stabilizing selection with theoretical motivation for form of stabilizing selection
%  - plot: (A) trait mean over time and trait distribution at the end
%  - plot: (B) (distrn of seg noise), (seg noise vs midparent), and (Radon-Nicodym deriv of conditional seg noise for a few values of midparent)
% 
% If we remove 'big loci' is what's left the infinitesimal model?
%  - people do this in practice (animal breeding papers)
%  - theory:  form of limiting conditional distribution
%  - theory: something about independence  (?? if we get there ??)
%  - plots (A) & (B) above from simulation, but for only small effects

%%%% LAST VERSION
% 1. *Intro*: Motivation about how many results about quant gen and methods assume Gaussian distributions.
%     Mention Levy walk macroevolution papers.
%     Point out that usual quant gen has *no* sweeps, while stable distributions would have sweeps (since some s would be bigger than 1/N).
%     Motivate with mixture model of effect sizes.
%     If stable distributions are a better model, what does that tell us? We can better understand the *noise* in evolution, i.e., predictability and saltation-ness.
%     (PETER)
% 
% 2. Motivation: empirical SNPs distributions in real data (PETER)
% 
% 3. Introducing infinitesimal model (in this case) (TODD)
% 
% 4. Proof following BEV in stable case (TODD)
% 
% 5. Theoretical results on stationary distributions (TODD)
% 
% 6. Simulations (PETER)

\title{\large{\bf
    Abnormal quantitative genetics?
}}

@@

Much of the work to give a quantitative understanding to the observation that
offspring of sexual organisms resemble their parent,
has focused on \emph{quantitative genetics},
which is interestingly intermediate between a mechanistic model of genetics
and a purely phenomenological model.
The action of natural selection on genetic variants in the absence of genetic variation
is relatively well-understood (cite ewens),
but tractable mathematical models of evolution for recombining species
are elusive,
at least with some results of when the model might be a good description of reality.
For instance, a great deal of conceptual understanding and even quantitative predictions
have been made assuming that the distribution of trait values in a population is Gaussian
(cite Lande etcetera)
despite the absence of a plausible model that would make this true (cite turelli and barton).
One of the oldest and most concrete models of heritable trait evolution is the \emph{infinitesimal model}
which assumes a simple, explicit form for how an offspring's trait is inherited from their parents':
it is Normally distributed, with mean equal to the average of the parents' values
and a variance independent of their values.
Despite the plausibility of this model,
(thanks to the Central Limit Theorem),
it was not until \citet{barton2017infinitesimal} that conditions under which this would be a good approximation
were made explicit,
through a formal convergence proof.

@@

Concretely, the infinitesimal model postulates that
if the parents' genetic contributions to the trait (i.e., their \emph{breeding values})
are $z^{[1]}$ and $z^{[2]}$,
and the offspring's genetic contribution to the trait is $z$,
and $R = z - (z^{[1]} + z^{[2]})/2$ is the deviation from the midparent
(which we call the \emph{segregation noise}), then
\begin{enumerate}
    \item $R$ is independent of $z^{[1]}$ and $z^{[2]}$, and
    \item $R$ has a Gaussian distribution.
\end{enumerate}
(This is not a full definition,
since the model makes broader statements involving conditioning on the full pedigree;
see \citet{barton2017infinitesimal}.)
We set out here to see whether (2) can be generalized,
and find that (1) is apparently inexitrably connected.

The Gaussian distribution appears thanks to ``the'' Central Limit Theorem,
i.e., the universality
of adding up a great many very small and independent things.
However, as is well known to probabilists,
there is a wider class of Central Limit Theorems,
which describe the distributions of sums of a great many independent and exchangeable things:
the \emph{stable laws}.
Roughly speaking, these say that if we add together $n$ independent copies of a random variable $X$
for which $\P\{ X > t \} \propto t^{-\alpha}$ for some $0 < \alpha < 2$,
then -- almost regardless of the details -- the result, scaled by $n^{1/\alpha}$,
is well-approximated by a universal form, the $\alpha$-stable distribution (and becomes exact, as $n \to \infty$).
In such ensembles, sinle large entries are important:
concretely, the largest of the $n$ copies will be of order $n^{1/\alpha}$,
so that e.g., for $\alpha=1$ a positive proportion of the sum of $n$ entries will be contributed by a single one.

This suggests exploring whether stable distributions might reasonably stand in for the Gaussian
in the infinitesimal model.
In this paper, we make preliminary explorations in that direction:
Does the universality shown by \citet{barton2017infinitesimal}
carry over when the distribution of effect sizes falls in the domain of attraction of a stable law?
What can be said about equilibrium trait distributions in this case?
And, is there any biological evidence that this theory might be relevant?

@@

%%%%%%%%%%%%%%%%%%%%%%
\section{Non-Gaussian effects}

We generally follow the notation of \citet{barton2017infinitesimal},
although in a simpler situation,
as we consider only the haploid offspring of a single set of haploid parents
rather than full pedigrees,
and ignore non-genetic contributions to the trait;
so, below we say ``trait'' to mean ``genetic contribution to the trait''.

Suppose that the focal individual's trait is
$$
    Z_M = \sum_{\ell=1}^M \eta_\ell / M^{1/\alpha},
$$
where $\eta_\ell$ is the contribution of the allele carried at the $\ell^\text{th}$ locus.
The effects of the alleles carried by the two parents
are $\eta_\ell^{[1]}$ and $\eta_\ell^{[2]}$, respectively,
and their traits are
$Z_M^{[1]} = \sum_\ell \eta_\ell^{[1]} / M^{1/\alpha}$ and
$Z_M^{[2]} = \sum_\ell \eta_\ell^{[2]} / M^{1/\alpha}$.
We assume the loci are unlinked and inheritance is fair, so
letting $X_\ell$ be a family of Bernoulli(1/2) random variables that are independent of everything else,
$$
    \eta_\ell = X_\ell \eta_\ell^{[1]} + (1 - X_\ell) \eta_\ell^{[1]} .
$$

We know from \citet{barton2017infinitesimal} that if the $\eta^{[i]}_\ell$ are
independent with mean zero and $\sum_{\ell=1}^M \var[\eta_\ell]/M \to \sigma^2$,
then with $\alpha=2$,
the traits $(Z_M, Z_M^{[1]}, Z_M^{[2]})$ converge in distribution to a multivariate Gaussian distribution
$(Z, Z^{[1]}, Z^{[2]})$ for which $Z - (Z^{[1]} + Z^{[2]})/2$ is independent of $Z^{[1]}$ and $Z^{[2]}$.

Now let us choose the distribution of allelic effects so that
the trait distributions converge to a nontrivial limit after rescaling by a different exponent $\alpha$.
Can the property of independence be preserved?
The following calculations suggest the answer is ``no``.

\begin{lem} \label{lem:nongaussian_nonindep}
    \comment{SOMETHING LIKE THE FOLLOWING?}
    Suppose that the allelic effects are such that
    $$
        \lim_{M \to \infty} \log \E[ \prod_{\ell=1}^M e^{it\eta^{[i]}} ]
        = c |t|^\alpha .
    $$
    Then $(Z_M, Z_M^{[1]}, Z_M^{[2]})$ converges in distribution
    to $(Z, Z^{[1]}, Z^{[2]})$ for which each of the three traits
    marginally has a standard $\alpha$-stable distribution,
    but the conditional distribution of $R = Z - (Z^{[1]} - Z^{[2]}))/2$
    given $Z^{[1]}=z_1$ and/or $Z^{[2]}=z_2$ depends nontrivially on $z_1$ and $z_2$.
\end{lem}

\begin{proof}[of Lemma~\ref{lem:nongaussian_nonindep}]
    \comment{calculations with characteristic functions?}
\end{proof}

This result should perhaps not be surprising:
there is a deep connection between independence and the Gaussian distribution.
For instance, a proof of the basic properties of the (Gaussian) infinitesimal model
might be sketched as follows:
divide the parental loci into groups according to the results of inheritance,
defining
\begin{align*}
    U_M^{[i]} = \sum_{\ell=1}^M X_\ell \eta_\ell^{[i]} 
    \qquad \text{ and } \qquad
    V_M^{[i]} = \sum_{\ell=1}^M (1-X_\ell) \eta_\ell^{[i]} ,
\end{align*}
so that 
$Z_M^{[1]} = U_M^{[1]} + V_M^{[1]}$ and
$Z_M^{[2]} = U_M^{[2]} + V_M^{[2]}$, so
\begin{align*}
    Z_M = U_M^{[1]} + V_M^{[2]} \qquad \text{ and } \qquad
    R_M = \frac{U_M^{[1]} - V_M^{[1]} - U_M^{[2]} + V_M^{[2]}}{ 2 } .
\end{align*}
Suppose we then can prove (as is indeed true in the Gaussian case) that
(a) $(U_M^{[1]}, V_M^{[1]}, U_M^{[2]}, V_M^{[2]})$ converge in distribution;
(b) the limiting random variables are independent; and
defining $R$, $Z^{[1]}$ and $Z^{[2]}$ as is natural in the limit,
(c) $R$ is independent of $Z^{[1]}$ and $Z^{[2]}$.
The Kac-Bernstein theorem states that if two independent random variables
have the property that if their sum and difference are independent of each other,
then the random variables have a Gaussian distribution
(\citet{kac1939characterization,bernstein1941property},
but see \citet{kagan1973characterization} for many much more general results).
It is not hard to see that (c) implies that $U^{[1]}$ and $V^{[1]}$
are such a pair of random variables, and therefore Gaussian.

Is there a converse? In other words,
does independence of segregation noise imply a Gaussian distribution of trait values?
It is not clear to us in how much generality such a statement might be made,
but we can at least prove the following:

\begin{lemma}\label{lem:parental_contribs}
    Under the conditions of Lemma~\ref{lem:nongaussian_nonindep},
    $U^{[1]}$ and $V^{[1]}$ are not independent.
    \comment{OR SOMETHING?}
\end{lemma}

\begin{proof}[of Lemma~\ref{lem:parental_contribs}]
    \comment{more calcs with characteristic functions?}
\end{proof}
